Let's talk about Thurstone \& Co.: An information-theoretical model for comparative judgments, and its statistical translation

### 2025-02-09 _tex/paper2.tex
Built site for gh-pages

In aggregate, the excessive reliance on Thurstone's Case V assumptions
in the statistical analysis of CJ data can compromise the reliability of
CJ's trait estimates. This overreliance, coupled with the apparent
disconnect between CJ's approach to trait measurement and hypothesis
testing, may also undermine the validity of these estimates
\citep[pp.~2]{Perron_et_al_2015}. Thus, addressing these issues through
a more systematic and integrated approach could provide significant
statistical and measurement benefits. For this reason, the study employs
the structural approach to causal inference, as it fulfills these
requirements.

\section{Extending Thurstone's model}\label{sec-theoretical}

@@

A motivating example helps to clarify these concepts. The example will
appear throughout the document to illustrate various aspects of the
proposed extension. In this scenario, researchers aim to determine: ``To
what extent do different teaching methods influence students' ability to
produce high-quality written texts?'' To investigate this, they could
design a CJ experiment by randomly assigning students (individuals) to
two groups, each receiving a different teaching method. Judges would
then compare pairs of students' written texts (stimuli), producing a
dichotomous outcome that reflects the relative quality of each text
(trait). Based on this setup, researchers could reformulate the research
question as the estimand: ``\emph{On average}, is there a difference in
the ability to produce high-quality written texts between the two groups
of students?''. Following current CJ practices, researchers would then
rely on estimates from the BTL model, or its transformations, to
approximate this estimand.

@@

As a result, this section addresses the identified limitations by
extending Thurstone's general form (see Table~\ref{tbl-thurstone_cases})
through a systematic and integrated approach based on the structural
approach to causal inference. Specifically, it leverages the
capabilities of this approach to formalize the combination of the core
theoretical principles outlined in Section~\ref{sec-thurstone_theory}
with key assessment design features relevant to CJ experiments, such as
the selection of judges, stimuli, and comparisons.

In addition to addressing the identified limitations, the extended model
offers two key advantages. First, it clarifies the interactions among
all actors and processes involved in CJ experiments. Second, it shifts
the current data analysis paradigm from passively accepting CJ's model
assumptions to actively testing them with the specific data under
analysis. For more details on these assumptions, refer to
Section~\ref{sec-theory-issues}.

@@

Before incorporating the theoretical principles into the
\emph{conceptual-population model}, the study defines SCMs, which are
formal mathematical models characterized by a set of \emph{endogenous}
variables \(V\), a set of \emph{exogenous} variables \(E\), and a set of
functions \(F\) \citep{Pearl_2009, Cinelli_et_al_2020}. Endogenous
variables are those whose causal mechanisms a researcher chooses to
model \citep{Neal_2020}. In contrast, exogenous variables represent
\emph{errors} or \emph{disturbances} arising from omitted factors that
the investigator chooses not to model explicitly
\citep[pp.~27,68]{Pearl_2009}. Lastly, the functions, referred to as
\emph{structural equations}, express the endogenous variables as
non-parametric functions of other endogenous and exogenous variables.
These functions use the symbol `\(:=\)' to denote the asymmetrical
causal dependence between variables and the symbol `\(\:\bot\:\)' to
represent \emph{d-separation}, a concept akin to (conditional)
independence.

@@

\[
\begin{aligned}
  O_{abhijk} & := f_{O}(D_{abhijk}) \\
  D_{abhijk} & := f_{D}(PT_{ahjk}, PT_{bijk}) \\
  PT_{ahjk} & := f_{PT}(T_{ah}, B_{jk}) \\
  PT_{bijk} & := f_{PT}(T_{bi}, B_{jk})
\end{aligned}
\]

@@

@@

\includegraphics[width=0.45\linewidth,height=\textheight,keepaspectratio]{./images/png/CJ_TM_04.png}

@@

@@

@@

\[
\begin{aligned}
  O_{PRQ} & := f_{O}(D_{PRQ}) \\
  D_{PRQ} & := f_{D}(PT_{PRQ}) \\
  PT_{PRQ} & := f_{PT}(T_{AI}, B_{KJ})
\end{aligned}
\]

@@

\subcaption{\label{fig-cj05_scm}SCM}

@@

\includegraphics[width=0.44\linewidth,height=\textheight,keepaspectratio]{./images/png/CJ_TM_05.png}

@@

\subcaption{\label{fig-cj05_dag}DAG}

@@

\caption{\label{fig-cj05}CJ's outcome, discriminal difference,
discriminal processes and judges' biases}

@@

\[
\begin{aligned}
  O_{PRQ} & := f_{O}(D_{PRQ}) \\
  D_{PRQ} & := f_{D}(PT_{PRQ}) \\
  PT_{PRQ} & := f_{PT}(T_{AI}, B_{KJ}) \\
  T_{AI} & := f_{T}(T_{I}, X_{A}, e_{A}) \\
  T_{I} & := f_{T}(X_{I}, e_{I}) \\
  B_{KJ} & := f_{B}(B_{J}, Z_{K}, e_{K}) \\
  B_{J} & := f_{B}(Z_{J}, e_{J}) 
\end{aligned}
\]

@@

\subcaption{\label{fig-cj10_scm}SCM}

@@

\includegraphics[width=0.68\linewidth,height=\textheight,keepaspectratio]{./images/png/CJ_TM_10.png}

@@

\subcaption{\label{fig-cj10_dag}DAG}

@@

\caption{\label{fig-cj10}CJ's outcome, discriminal difference,
discriminal processes and judges' biases}

@@

\[
\begin{aligned}
  O^{S}_{PRQ} & := f_{O}(O_{PRQ}, S_{I}, S_{A}, S_{K}, S_{J}) \\
  O_{PRQ} & := f_{O}(D_{PRQ}) \\
  D_{PRQ} & := f_{D}(PT_{PRQ}) \\
  PT_{PRQ} & := f_{PT}(T_{AI}, B_{KJ}) \\
  T_{AI} & := f_{T}(T_{I}, X_{A}, e_{A}) \\
  T_{I} & := f_{T}(X_{I}, e_{I}) \\
  B_{KJ} & := f_{B}(B_{J}, Z_{K}, e_{K}) \\
  B_{J} & := f_{B}(Z_{J}, e_{J})
\end{aligned}
\]

@@

\subcaption{\label{fig-cj12_scm}SCM}

@@

\includegraphics[width=0.8\linewidth,height=\textheight,keepaspectratio]{./images/png/CJ_TM_12.png}

@@

\subcaption{\label{fig-cj12_dag}DAG}

@@

\caption{\label{fig-cj12}CJ's outcome, discriminal difference,
discriminal processes and judges' biases}

@@

\[
\begin{aligned}
  O^{C}_{PRQ} & := f_{C}(O^{S}_{PRQ}, C) \\
  O^{S}_{PRQ} & := f_{S}(O_{PRQ}, S_{I}, S_{A}, S_{K}, S_{J}) \\
  O_{PRQ} & := f_{O}(D_{PRQ}) \\
  D_{PRQ} & := f_{D}(PT_{PRQ}) \\
  PT_{PRQ} & := f_{PT}(T_{AI}, B_{KJ}) \\
  T_{AI} & := f_{T}(T_{I}, X_{A}, e_{A}) \\
  T_{I} & := f_{T}(X_{I}, e_{I}) \\
  B_{KJ} & := f_{B}(B_{J}, Z_{K}, e_{K}) \\
  B_{J} & := f_{B}(Z_{J}, e_{J}) 
\end{aligned}
\]

@@

\subcaption{\label{fig-cj14_scm}SCM}

@@

\includegraphics[width=0.9\linewidth,height=\textheight,keepaspectratio]{./images/png/CJ_TM_14.png}

@@

\subcaption{\label{fig-cj14_dag}DAG}

@@

\caption{\label{fig-cj14}CJ's outcome, discriminal difference,
discriminal processes and judges' biases}

\[
\begin{aligned}
  O^{C}_{PRQ} & := f_{C}(O^{S}_{PRQ}, C) \\
  O^{S}_{PRQ} & := f_{S}(O_{PRQ}, S_{I}, S_{A}, S_{K}, S_{J}) \\
  O_{PRQ} & := f_{O}(D_{PRQ}) \\
  D_{PRQ} & := f_{D}(PT_{PRQ}) \\
  PT_{PRQ} & := f_{PT}(T_{AI}, B_{KJ}) \\
  T_{AI} & := f_{T}(T_{I}, X_{A}, e_{A}) \\
  T_{I} & := f_{T}(X_{I}, e_{I}) \\
  B_{KJ} & := f_{B}(B_{J}, Z_{K}, e_{K}) \\
  B_{J} & := f_{B}(Z_{J}, e_{J}) 
\end{aligned}
\]

\subcaption{\label{fig-cj15_scm}SCM}

\includegraphics[width=0.9\linewidth,height=\textheight,keepaspectratio]{./images/png/CJ_TM_15.png}

\subcaption{\label{fig-cj15_dag}DAG}

\caption{\label{fig-cj15}CJ's outcome, discriminal difference,
discriminal processes and judges' biases}

@@

\textbf{Funding:} The project was founded through the Research Fund of
the University of Antwerp (BOF).

\textbf{Financial interests:} The authors have no relevant financial
interest to disclose.

\textbf{Non-financial interests:} The authors have no relevant
non-financial interest to disclose.

\textbf{Ethics approval:} The University of Antwerp Research Ethics
Committee has confirmed that no ethical approval is required.

@@

\textbf{Consent for publication:} All authors have read and agreed to
the published version of the manuscript.

\textbf{Availability of data and materials:} No data was utilized in
this study.

\textbf{Code availability:} All the code utilized in this research is
available in the digital document located at:
\url{https://jriveraespejo.github.io/paper2_manuscript/}.

\textbf{AI-assisted technologies in the writing process:} The authors
utilized a range of AI-based language tools throughout the preparation
of this work. They occasionally employed the tools to refine phrasing
and optimize wording, ensuring appropriate language use and enhancing
the manuscript's clarity and coherence. The authors take full
responsibility for the final content of the publication.

@@

@@

\subsection{Empirical research and randomized
experiments}\label{sec-appendix-A}

@@

\subsection{Identification under causal inference}\label{sec-appendix-B}

@@

\emph{Counterfactuals} are hypothetical scenarios that are
\emph{contrary to fact}, where alternative outcomes resulting from a
given cause are neither observed nor observable
\citep{Neal_2020, Counterfactual_2024}. The structural approach to
causal inference \citep{Pearl_2009, Pearl_et_al_2016} provides a formal
framework for defining counterfactuals. For instance, in the scenario
described in Section~\ref{sec-appendix-A}, the approach begins by
defining the \emph{individual causal effect} (ICE) as the difference
between each student's potential outcomes, as in Equation~\ref{eq-ICE}.

@@

For instance, if researchers cannot conduct the randomized experiments
described in Section~\ref{sec-appendix-A} and must instead rely on
observational data, they can still identify the ACE as long as an
observed variable \(X\), such as the socio-economic status of the
school, satisfies the backdoor criterion. Under these circumstances,
researchers first identify the \emph{conditional average causal effect}
(CACE, Equation~\ref{eq-CACE})

@@

\subsubsection{Diving into the specifics}\label{sec-appendix-B2}

@@

SCMs and DAGs support identification analysis through two key
advantages. First, regardless of complexity, they can represent various
causal structures using only five fundamental building blocks
\citep{Neal_2020, McElreath_2020}. This feature allows researchers to
decompose complex structures into manageable components, facilitating
their analysis \citep{McElreath_2020}. Second, they depict causal
relationships in a non-parametric and fully interactive way. This
flexibility enables feasible ACE identification strategies without
defining the variables' data types, the functional form between them, or
their parameters \citep[pp.~35]{Pearl_et_al_2016}.
Section~\ref{sec-appendix-B21} and Section~\ref{sec-appendix-B22}
elaborate on the first advantage, while Section~\ref{sec-appendix-B22}
and Section~\ref{sec-appendix-B23} do so for the second.

\paragraph{The five fundamental block for SCMs and
DAGs}\label{sec-appendix-B21}

@@

Researchers can then use these building blocks to represent the scenario
outlined in Section~\ref{sec-appendix-B}. SCM \ref{fig-scm_example1} and
DAG \ref{fig-mdag_example1} depict the plausible causal structure for
this example. In this context, the variable \(X\) (socio-economic status
of the school) is thought to be a confounder in the relationship between
the teaching method \(T\) and the outcome \(Y\). The figures display
multiple descendant relationships such as \(X \rightarrow T\),
\(X \rightarrow Y\), and \(T \rightarrow Y\). They also highlight
unconnected node pairs, evident from the relationships
\(e_{T} \:\bot\:e_{X}\), \(e_{T} \:\bot\:e_{Y}\), and
\(e_{X} \:\bot\:e_{Y}\). Additional, the figures show one fork,
\(X \rightarrow \{T, Y\}\), and two colliders:
\(\{X, e_{T}\} \rightarrow T\) and \(\{X, T, e_{Y}\} \rightarrow Y\).

@@

\caption{\label{fig-example1}Plausible causal structure the scenario
outlined in Section~\ref{sec-appendix-B}.}

@@

\paragraph{The probabilistic implications of these
blocks}\label{sec-appendix-B22}

@@

The five fundamental building blocks described in
Section~\ref{sec-appendix-B21} clearly illustrate which (in)dependencies
can SMCs and DAGs predict. For instance, applying the BNF to the causal
structure shown in the SCM \ref{fig-scm_bb1} and DAG \ref{fig-mdag_bb1}
enables researchers to express the joint probability distribution of the
observed variables as \(P(X_{1}, X_{3}) = P(X_{1}) P(X_{3})\),
supporting the theoretical assumption that the observed variables
\(X_{1}\) and \(X_{3}\) are unconditionally independent
(\(X_{1} \:\bot\:X_{3}\)) \citep[pp.~24]{Neal_2020}. Conversely, when
\(X_{3}\) is unconditionally dependent on \(X_{1}\)
(\(X_{1} \:\not\bot\:X_{3}\)), as depicted in the SCM \ref{fig-scm_bb2}
and DAG \ref{fig-mdag_bb2}, the BNF express their joint probability
distribution as \(P(X_{1}, X_{3}) = P(X_{3} \:|\:X_{1}) P(X_{1})\).
Notably, these descriptions demonstrate the clear correspondence between
the structural equations illustrated in Section~\ref{sec-appendix-B21}
and the CPDs.

@@

Using these additional probabilistic insights, researchers can revisit
the scenario in Section~\ref{sec-appendix-B}. In this context, applying
the BNF to the SCM \ref{fig-scm_example2} structure, enables the
representation of the joint probability distribution of the observed
variables as \(P(Y, T, X) =\) \(P(Y \:|\:T, X) P(T \:|\:X) P(X)\). From
this expression, researchers can infer that the outcome \(Y\) is
unconditionally dependent on the teaching method \(T\)
\((Y \:\not\bot\:T)\). This dependency arises from two key structures: a
direct causal path from the teaching method \(T\) to the outcome \(Y\),
represented by the two-connected-nodes structure \(T \rightarrow Y\)
(black path in DAG \ref{fig-mdag_example2}), and a confounding
non-causal path from the teaching method \(T\) to the outcome \(Y\)
through the socio-economic status of the school \(X\), represented by
the fork structure \(T \leftarrow X \rightarrow Y\) (gray path in DAG
\ref{fig-mdag_example2}).

@@

\caption{\label{fig-example2}Plausible causal structure the scenario
outlined in Section~\ref{sec-appendix-B}.}

@@

\paragraph{From probability to causality}\label{sec-appendix-B23}

@@

With these new insights, researchers revisiting the scenario in
Section~\ref{sec-appendix-B22} can infer that the socio-economic status
of the school, \(X\), satisfies the backdoor criterion, assuming the
causal structure depicted by the SCM \ref{fig-scm_example2} and DAG
\ref{fig-mdag_example2} is correct. This means that \(X\) serves as a
sufficient adjustment set, as it effectively blocks all confounding
non-causal paths introduced by the fork structure. Nevertheless, since
\(Y\) remains dependent on \(T\) even after conditioning
\((Y \:\not\bot\:T \:|\:X)\), this dependency can only be attributed to
the direct causal effect \(T \rightarrow Y\). Notably, for the purpose
of identification, the conditioned DAG \ref{fig-mdag_example2} is
equivalent to the manipulated DAG \ref{fig-mdag_example3}, because \(X\)
satisfies the backdoor criterion.

@@

\caption{\label{fig-example3}Plausible causal structure the scenario
outlined in Section~\ref{sec-appendix-B22}.}

@@

Researchers can then apply the \emph{backdoor adjustment} to identify
the ACE of \(T\) on \(Y\). They achieve this by first identifying the
CACE of \(T\) on \(Y\) by conditioning on \(X\), and then marginalizing
this effect over \(X\) to obtain the ACE. This process is expressed in
Equation~\ref{eq-mACE2} (see Section~\ref{sec-appendix-B}).

@@

\paragraph{The estimation process}\label{sec-appendix-C}

@@

Nevertheless, researchers can manage the complexity of the likelihood by
assuming specific local (in)dependencies among variables. SCMs and DAGs
provide a formal framework to represent these assumptions, as detailed
in Section~\ref{sec-appendix-B22}. These assumptions improve model
tractability and simplify the estimation process by enabling the
derivation of the BNF of the likelihood (Equation~\ref{eq-like_BNF}).
With this simplified structure, any probabilistic programming language
can model the system and compute the parameter's posterior distribution
using Equation~\ref{eq-bayes_rule}.

>>>>

In aggregate, the excessive reliance on Thurstone's Case V assumptions
in the statistical analysis of CJ data can compromise the reliability of
CJ's trait estimates. This overreliance, coupled with the apparent
disconnect between CJ's approach to trait measurement and hypothesis
testing, may also undermine the validity of these estimates
\citep[pp.~2]{Perron_et_al_2015}. Thus, addressing these issues through
a more systematic and integrated approach could provide significant
statistical and measurement benefits. For this reason, the study employs
the structural approach to causal inference.

\section{Extending Thurstone's general form}\label{sec-theoretical}

@@

A motivating example helps to clarify these concepts. This example will
appear throughout the document to illustrate various aspects of the
proposed extension. In this scenario, researchers aim to determine: ``To
what extent do different teaching methods influence students' ability to
produce high-quality written texts?'' To investigate this, they design a
CJ experiment by randomly assigning students (individuals) to two
groups, each receiving a different teaching method. Judges then compare
pairs of students' written texts (stimuli), producing a dichotomous
outcome that reflects the relative quality of each text (trait). Based
on this setup, researchers could reformulate the research question as
the estimand: ``\emph{On average}, is there a difference in the ability
to produce high-quality written texts between the two groups of
students?''. Following current CJ practices, researchers would then rely
on estimates from the BTL model, or its transformations, to approximate
this estimand.

@@

Thus, this section addresses the identified limitations by extending
Thurstone's general form (see Table~\ref{tbl-thurstone_cases}) through a
systematic and integrated approach based on the structural approach to
causal inference. Specifically, it leverages the capabilities of this
approach to formalize the combination of the core theoretical principles
outlined in Section~\ref{sec-thurstone_theory} with key assessment
design features relevant to CJ experiments, such as the selection of
judges, stimuli, and comparisons.

In addition to overcoming the identified limitations, the extended model
offers two key advantages. First, it clarifies the interactions among
all actors and processes involved in CJ experiments. Second, it shifts
the current comparative data analysis paradigm from passively accepting
the model assumptions to actively testing if those assumptions fit the
specific data under analysis. For more details on these assumptions,
refer to Section~\ref{sec-theory-issues}.

@@

Before incorporating the theoretical principles into the
\emph{conceptual-population model}, the study defines SCMs. SCMs are
formal mathematical models characterized by a set of \emph{endogenous}
variables \(V\), a set of \emph{exogenous} variables \(E\), and a set of
functions \(F\) \citep{Pearl_2009, Cinelli_et_al_2020}. Endogenous
variables are those whose causal mechanisms a researcher chooses to
model \citep{Neal_2020}. In contrast, exogenous variables represent
\emph{errors} or \emph{disturbances} arising from omitted factors that
the investigator chooses not to model explicitly
\citep[pp.~27,68]{Pearl_2009}. Lastly, the functions, referred to as
\emph{structural equations}, express the endogenous variables as
non-parametric functions of other endogenous and exogenous variables.
These functions use the symbol `\(:=\)' to denote the asymmetrical
causal dependence between variables and the symbol `\(\:\bot\:\)' to
represent \emph{d-separation}, a concept akin to (conditional)
independence.

@@

\[
\begin{aligned}
  O^{cp}_{abhijk} & := f_{O}(D_{abhijk}) \\
  D_{abhijk} & := f_{D}(T_{ah}, T_{bi}, B_{jk})
\end{aligned}
\]

@@

\subcaption{\label{fig-cj03_scm}SCM}

@@

\includegraphics[width=0.35\linewidth,height=\textheight,keepaspectratio]{./images/png/CJ_TM_03.png}

@@

\subcaption{\label{fig-cj03_dag}DAG}

@@

\caption{\label{fig-cj03}CJ's outcome, discriminal difference,
discriminal processes and judges' biases}

@@

\[
\begin{aligned}
  O^{cp}_{RPUV} & := f_{O}(D_{RPUV}) \\
  D_{RPUV} & := f_{D}(T_{IA}, B_{JK})
\end{aligned}
\]

@@

@@

\includegraphics[width=0.44\linewidth,height=\textheight,keepaspectratio]{./images/png/CJ_TM_04.png}

@@

@@

@@

\[
\begin{aligned}
  O^{cp}_{RPUV} & := f_{O}(D_{RPUV}) \\
  D_{RPUV} & := f_{D}(T_{IA}, B_{JK}) \\
  T_{IA} & := f_{T}(T_{I}, X_{A}, e_{IA}) \\
  T_{I} & := f_{T}(X_{I}, e_{I}) \\
  B_{JK} & := f_{B}(B_{J}, Z_{K}, e_{JK}) \\
  B_{J} & := f_{B}(Z_{J}, e_{J}) 
\end{aligned}
\]

@@

\subcaption{\label{fig-cj09_scm}SCM}

@@

\includegraphics[width=0.7\linewidth,height=\textheight,keepaspectratio]{./images/png/CJ_TM_09.png}

@@

\subcaption{\label{fig-cj09_dag}DAG}

@@

\caption{\label{fig-cj09}CJ's outcome, discriminal difference,
discriminal processes and judges' biases}

@@

\[
\begin{aligned}
  O^{sc}_{RPUV} & := f_{O}(O^{cp}_{RPUV}, S_{I}, S_{A}, S_{K}, S_{J}) \\
  O^{cp}_{RPUV} & := f_{O}(D_{RPUV}) \\
  D_{RPUV} & := f_{D}(T_{IA}, B_{JK}) \\
  T_{IA} & := f_{T}(T_{I}, X_{A}, e_{IA}) \\
  T_{I} & := f_{T}(X_{I}, e_{I}) \\
  B_{JK} & := f_{B}(B_{J}, Z_{K}, e_{JK}) \\
  B_{J} & := f_{B}(Z_{J}, e_{J}) 
\end{aligned}
\]

@@

\subcaption{\label{fig-cj11_scm}SCM}

@@

\includegraphics[width=0.84\linewidth,height=\textheight,keepaspectratio]{./images/png/CJ_TM_11.png}

@@

\subcaption{\label{fig-cj11_dag}DAG}

@@

\caption{\label{fig-cj11}CJ's outcome, discriminal difference,
discriminal processes and judges' biases}

@@

\[
\begin{aligned}
  O_{RPUV} & := f_{C}(O^{sc}_{RPUV}, C) \\
  O^{sc}_{RPUV} & := f_{O}(O^{cp}_{RPUV}, S_{I}, S_{A}, S_{K}, S_{J}) \\
  O^{cp}_{RPUV} & := f_{O}(D_{RPUV}) \\
  D_{RPUV} & := f_{D}(T_{IA}, B_{JK}) \\
  T_{IA} & := f_{T}(T_{I}, X_{A}, e_{IA}) \\
  T_{I} & := f_{T}(X_{I}, e_{I}) \\
  B_{JK} & := f_{B}(B_{J}, Z_{K}, e_{JK}) \\
  B_{J} & := f_{B}(Z_{J}, e_{J}) 
\end{aligned}
\]

@@

\subcaption{\label{fig-cj13_scm}SCM}

@@

\includegraphics[width=1\linewidth,height=\textheight,keepaspectratio]{./images/png/CJ_TM_13.png}

@@

\subcaption{\label{fig-cj13_dag}DAG}

@@

\caption{\label{fig-cj13}CJ's outcome, discriminal difference,
discriminal processes and judges' biases}

@@

\textbf{Funding:} The Research Fund (BOF) of the University of Antwerp
funded this project.

\textbf{Financial interests:} The authors declare no relevant financial
interests.

\textbf{Non-financial interests:} The authors declare no relevant
non-financial interests.

\textbf{Ethics approval:} The University of Antwerp Research Ethics
Committee confirmed that this study does not require ethical approval.

@@

\textbf{Consent for publication:} All authors have read and approved the
final version of the manuscript for publication.

\textbf{Data availability:} This study did not use any data.

\textbf{Materials and code availability:} The \texttt{CODE\ LINK}
section at the top of the digital document located at:
\url{https://jriveraespejo.github.io/paper2_manuscript/} provides access
to all materials and code.

\textbf{AI-assisted technologies in the writing process:} The authors
used various AI-based language tools to refine phrasing, optimize
wording, and enhance clarity and coherence throughout the manuscript.
They take full responsibility for the final content of the publication.

@@

\subsection{Statistics and Causal inference}\label{sec-appendixB}

@@

\subsubsection{Empirical research and randomized
experiments}\label{sec-appendixB1}

@@

\subsubsection{Identification under causal
inference}\label{sec-appendixB2}

@@

\emph{Counterfactuals} are hypothetical scenarios that are
\emph{contrary to fact}, where alternative outcomes resulting from a
given cause are neither observed nor observable
\citep{Neal_2020, Counterfactual_2024}. The structural approach to
causal inference \citep{Pearl_2009, Pearl_et_al_2016} provides a formal
framework for defining counterfactuals. For instance, in the scenario
described in Section~\ref{sec-appendixB1}, the approach begins by
defining the \emph{individual causal effect} (ICE) as the difference
between each student's potential outcomes, as in Equation~\ref{eq-ICE}.

@@

For instance, if researchers cannot conduct the randomized experiments
described in Section~\ref{sec-appendixB1} and must instead rely on
observational data, they can still identify the ACE as long as an
observed variable \(X\), such as the socio-economic status of the
school, satisfies the backdoor criterion. Under these circumstances,
researchers first identify the \emph{conditional average causal effect}
(CACE, Equation~\ref{eq-CACE})

@@

\subsubsection{Diving into the specifics}\label{sec-appendixB3}

@@

SCMs and DAGs support identification analysis through two key
advantages. First, regardless of complexity, they can represent various
causal structures using only five fundamental building blocks
\citep{Neal_2020, McElreath_2020}. This feature allows researchers to
decompose complex structures into manageable components, facilitating
their analysis \citep{McElreath_2020}. Second, they depict causal
relationships in a non-parametric and fully interactive way. This
flexibility enables feasible ACE identification strategies without
defining the variables' data types, the functional form between them, or
their parameters \citep[pp.~35]{Pearl_et_al_2016}.

Thus, Section~\ref{sec-appendixB31} and Section~\ref{sec-appendixB32}
elaborate on the first advantage, while Section~\ref{sec-appendixB32}
and Section~\ref{sec-appendixB33} do so for the second. Finally,
Section~\ref{sec-appendixB34} explains how researchers use SCMs and DAGs
alongside Bayesian inference methods in the estimation process.

\paragraph{The five fundamental block for SCMs and
DAGs}\label{sec-appendixB31}

@@

Researchers can then use these building blocks to represent the scenario
outlined in Section~\ref{sec-appendixB2}. SCM \ref{fig-scm_example1} and
DAG \ref{fig-mdag_example1} depict the plausible causal structure for
this example. In this context, the variable \(X\) (socio-economic status
of the school) is thought to be a confounder in the relationship between
the teaching method \(T\) and the outcome \(Y\). The figures display
multiple descendant relationships such as \(X \rightarrow T\),
\(X \rightarrow Y\), and \(T \rightarrow Y\). They also highlight
unconnected node pairs, evident from the relationships
\(e_{T} \:\bot\:e_{X}\), \(e_{T} \:\bot\:e_{Y}\), and
\(e_{X} \:\bot\:e_{Y}\). Additional, the figures show one fork,
\(X \rightarrow \{T, Y\}\), and two colliders:
\(\{X, e_{T}\} \rightarrow T\) and \(\{X, T, e_{Y}\} \rightarrow Y\).

@@

\caption{\label{fig-example1}Plausible causal structure the scenario
outlined in Section~\ref{sec-appendixB2}.}

@@

\paragraph{The probabilistic implications of these
blocks}\label{sec-appendixB32}

@@

The five fundamental building blocks described in
Section~\ref{sec-appendixB31} clearly illustrate which (in)dependencies
can SMCs and DAGs predict. For instance, applying the BNF to the causal
structure shown in the SCM \ref{fig-scm_bb1} and DAG \ref{fig-mdag_bb1}
enables researchers to express the joint probability distribution of the
observed variables as \(P(X_{1}, X_{3}) = P(X_{1}) P(X_{3})\),
supporting the theoretical assumption that the observed variables
\(X_{1}\) and \(X_{3}\) are unconditionally independent
(\(X_{1} \:\bot\:X_{3}\)) \citep[pp.~24]{Neal_2020}. Conversely, when
\(X_{3}\) is unconditionally dependent on \(X_{1}\)
(\(X_{1} \:\not\bot\:X_{3}\)), as depicted in the SCM \ref{fig-scm_bb2}
and DAG \ref{fig-mdag_bb2}, the BNF express their joint probability
distribution as \(P(X_{1}, X_{3}) = P(X_{3} \:|\:X_{1}) P(X_{1})\).
Notably, these descriptions demonstrate the clear correspondence between
the structural equations illustrated in Section~\ref{sec-appendixB31}
and the CPDs.

@@

Using these additional probabilistic insights, researchers can revisit
the scenario in Section~\ref{sec-appendixB2}. In this context, applying
the BNF to the SCM \ref{fig-scm_example2} structure, enables the
representation of the joint probability distribution of the observed
variables as \(P(Y, T, X) =\) \(P(Y \:|\:T, X) P(T \:|\:X) P(X)\). From
this expression, researchers can infer that the outcome \(Y\) is
unconditionally dependent on the teaching method \(T\)
\((Y \:\not\bot\:T)\). This dependency arises from two key structures: a
direct causal path from the teaching method \(T\) to the outcome \(Y\),
represented by the two-connected-nodes structure \(T \rightarrow Y\)
(black path in DAG \ref{fig-mdag_example2}), and a confounding
non-causal path from the teaching method \(T\) to the outcome \(Y\)
through the socio-economic status of the school \(X\), represented by
the fork structure \(T \leftarrow X \rightarrow Y\) (gray path in DAG
\ref{fig-mdag_example2}).

@@

\caption{\label{fig-example2}Plausible causal structure the scenario
outlined in Section~\ref{sec-appendixB2}.}

@@

\paragraph{From probability to causality}\label{sec-appendixB33}

@@

With these new insights, researchers revisiting the scenario in
Section~\ref{sec-appendixB32} can infer that the socio-economic status
of the school, \(X\), satisfies the backdoor criterion, assuming the
causal structure depicted by the SCM \ref{fig-scm_example2} and DAG
\ref{fig-mdag_example2} is correct. This means that \(X\) serves as a
sufficient adjustment set, as it effectively blocks all confounding
non-causal paths introduced by the fork structure. Nevertheless, since
\(Y\) remains dependent on \(T\) even after conditioning
\((Y \:\not\bot\:T \:|\:X)\), this dependency can only be attributed to
the direct causal effect \(T \rightarrow Y\). Notably, for the purpose
of identification, the conditioned DAG \ref{fig-mdag_example2} is
equivalent to the manipulated DAG \ref{fig-mdag_example3}, because \(X\)
satisfies the backdoor criterion.

@@

\caption{\label{fig-example3}Plausible causal structure the scenario
outlined in Section~\ref{sec-appendixB32}.}

@@

Researchers can then apply the \emph{backdoor adjustment} to identify
the ACE of \(T\) on \(Y\). They achieve this by first identifying the
CACE of \(T\) on \(Y\) by conditioning on \(X\), and then marginalizing
this effect over \(X\) to obtain the ACE. This process is expressed in
Equation~\ref{eq-mACE2} (see Section~\ref{sec-appendixB2}).

@@

\paragraph{The estimation process}\label{sec-appendixB34}

@@

Nevertheless, researchers can manage the complexity of the likelihood by
assuming specific local (in)dependencies among variables. SCMs and DAGs
provide a formal framework to represent these assumptions, as detailed
in Section~\ref{sec-appendixB32}. These assumptions improve model
tractability and simplify the estimation process by enabling the
derivation of the BNF of the likelihood (Equation~\ref{eq-like_BNF}).
With this simplified structure, any probabilistic programming language
can model the system and compute the parameter's posterior distribution
using Equation~\ref{eq-bayes_rule}.
